% ch11 — 3+1 ADM Decomposition
\chapter{3+1 ADM Decomposition}
\label{ch:adm-decomposition}

Einstein's field equations (\cref{sec:gr-efe}) are beautiful, compact, and
completely useless for direct numerical integration.  They are covariant ---
they treat space and time symmetrically --- but a computer needs something
different: data on a spatial surface at one instant and a rule for marching
that data forward to the next instant.  The Arnowitt--Deser--Misner (ADM)
formalism provides exactly this bridge.  It decomposes the four-dimensional
spacetime manifold into a one-parameter family of three-dimensional spatial
slices $\Sigma_t$, each carrying a Riemannian metric $\spatialmetric$ and
an extrinsic curvature $\extrinsic$ that together constitute the
gravitational initial data.  Two gauge fields --- the lapse $\lapse$ and
the shift $\shift$ --- encode the freedom in how coordinates are threaded
through the foliation, and Einstein's equations split into constraint
equations that the initial data must satisfy and evolution equations that
propagate the data forward.

The decomposition introduces five geometric objects that recur throughout
the rest of the book: the time function whose level sets define the slices,
the unit normal $n^\mu$ that threads them together, the spatial projection
operator $\gamma^\mu{}_\nu$ that separates tangent and normal components,
the induced metric $\gamma_{\mu\nu}$ that measures distances within each
slice, and the extrinsic curvature $K_{\mu\nu}$ that measures how each
slice bends within the ambient spacetime.  The chapter develops these
objects in sequence, starting from the foliation and building toward the
full ADM evolution system.  The final section shows that the ADM system as
written is only weakly hyperbolic --- a subtle but fatal deficiency for
numerical work that motivates the BSSN and CCZ4 reformulations of
\cref{ch:bssn-ccz4}.

% ---- Drafted sections ----
\input{part4_numerical_relativity/ch11_adm_decomposition/adm-foliation}

\input{part4_numerical_relativity/ch11_adm_decomposition/adm-lapse-shift}

\input{part4_numerical_relativity/ch11_adm_decomposition/adm-extrinsic-curvature}

\input{part4_numerical_relativity/ch11_adm_decomposition/adm-equations}

\paragraph{Key References.}
The ADM formalism follows \cite{ADM1962} and the pedagogical treatment in
\cite{Baumgarte2010}; the hyperbolicity discussion draws on
\cite{Alcubierre2008}.
