% ch05 — The Kerr Solution
\chapter{The Kerr Solution}
\label{ch:kerr}

The Schwarzschild metric of \cref{ch:schwarzschild} describes a non-rotating
black hole --- a useful idealisation, but not a realistic one.  X-ray binary
observations and Event Horizon Telescope imaging consistently find
dimensionless spins $a_* = a/M$ in the range $0.5$--$0.998$; a non-spinning
black hole is the exception, not the rule.  Rotation introduces qualitative
changes that go beyond the incremental: the single event horizon splits into
an outer and inner pair, frame-dragging creates an ergosphere where no
stationary observer can exist, and the innermost stable circular orbit
migrates from $6M$ to as close as $M$ for a maximally spinning prograde disc.
Each of these features shapes the images that the ray tracer produces.

Two coordinate systems carry the analysis.  Boyer--Lindquist coordinates
generalise the Schwarzschild $(t, r, \theta, \varphi)$ chart and expose the
analytic structure: the horizon function $\Delta = r^2 - 2Mr + a^2$ vanishes
at the horizons, the angular metric reveals frame-dragging, and the ISCO
formulae take closed form.  But $\Delta = 0$ is a coordinate singularity, just
as $r = 2M$ was for Schwarzschild --- and the geodesic integrator needs a
chart that stays finite at the horizon.  Kerr--Schild coordinates provide
this, decomposing the full Kerr metric as flat Minkowski space plus a
rank-one null correction, exactly as in the Schwarzschild case of
\cref{sec:schw-kerr-schild}.  The price is a quartic equation relating the
Boyer--Lindquist radius $r$ to the Cartesian coordinates $(x, y, z)$ --- a
relationship that must be inverted at every point where the integrator needs
to evaluate horizon proximity or disc geometry.

The chapter closes with two applications.  The ISCO radius and radiative
efficiency of thin accretion discs depend sensitively on spin: prograde
orbits around a rapidly spinning hole radiate over seven times more energy
per unit rest mass than orbits around a Schwarzschild hole, a difference
visible in both the brightness and geometry of the disc image.  A starred
section then examines the Morris--Thorne wormhole --- a non-black-hole
spacetime whose absence of horizons, nontrivial topology, and double-sky
images exercise code paths that no black hole touches.  The geodesic
machinery developed here feeds directly into \cref{ch:geodesics}, which
treats the numerical integration strategy, conservation monitoring, and
termination conditions that turn these metrics into rendered images.

% ---- Drafted sections ----
\input{part2_black_holes/ch05_kerr/kerr-bl}

\input{part2_black_holes/ch05_kerr/kerr-ks}

\input{part2_black_holes/ch05_kerr/kerr-bl-quartic}

\input{part2_black_holes/ch05_kerr/kerr-horizons}

\input{part2_black_holes/ch05_kerr/kerr-isco}

\input{part2_black_holes/ch05_kerr/kerr-wormhole}

\paragraph{Key References.}
The Kerr metric treatment follows \cite{Visser2007} and \cite{Wald1984}; ISCO
formulae are from \cite{Bardeen1972}; the wormhole section draws on
\cite{MorrisThorne1988}.
