% Appendix C — The event-horizon Codebase
\chapter{The \texttt{event-horizon} Codebase}
\label{app:codebase}

This appendix provides a map of the companion codebase, its architecture, and
instructions for building and running the software. The codebase is the
executable companion to the book: every algorithm derived in the text has a
corresponding implementation.

\section{Architecture: Haskell, Rust, and WGSL}
\label{sec:code-architecture}

The codebase is a multi-language project. Haskell provides the high-level
orchestration, mathematical modelling, and ray-tracing pipeline. Rust
implements performance-critical numerical kernels (evolution, solvers, MHD).
WGSL compute shaders run the GPU-accelerated geodesic integrator and
post-processing pipeline via wgpu.

\section{Module Map}
\label{sec:code-modules}

% TODO: Generate a comprehensive module listing as the codebase evolves.
A tabular module map will be provided here, listing each major module with its
language, purpose, and the chapter(s) that discuss it. Margin notes throughout
the book use \verb|\coderef{module}{symbol}| to link prose to source.

\section{Building and Running}
\label{sec:code-build}

We provide step-by-step instructions for building the project with Cabal
(Haskell), Cargo (Rust), and the wgpu shader pipeline, including the
required toolchain versions and platform-specific notes.

\section{Contributing}
\label{sec:code-contributing}

We outline the contribution workflow: branch conventions, code style, test
expectations, and the CI pipeline that validates every pull request.

\paragraph{Key References.}
The project repository is hosted at the URL given in the front matter.
Build toolchains: GHC \cite{GHC}, Cargo \cite{Cargo}, wgpu \cite{wgpu}.
