% ch10 — Spectral Methods
\chapter{Spectral Methods}
\label{ch:spectral-methods}

The finite difference stencils of \cref{ch:finite-differences} approximate
derivatives by combining a handful of neighbouring grid values --- five points
for a fourth-order first derivative, seven for the dissipation operator.
Adding more points to the stencil raises the order of accuracy, but the
improvement is only algebraic: doubling the stencil width roughly doubles the
convergence rate.  Spectral methods take the idea to its logical conclusion
by using \emph{every} point in the element to approximate the derivative at
each node.  The resulting differentiation matrix is dense rather than banded,
and for smooth functions the convergence is no longer algebraic but
\emph{exponential} --- each additional node can buy an extra digit of
accuracy rather than a fixed fraction of one.  This is the accuracy advantage
that makes high-order discontinuous Galerkin (DG) methods competitive with
finite differences at a fraction of the grid resolution.

The practical question is which polynomials and which nodes.  The chapter
opens with the interpolation theory that governs the choice, showing how
equidistant nodes fail catastrophically at high order (the Runge phenomenon)
and how clustered nodes cure the instability.  Two polynomial families
provide the clustering: Chebyshev polynomials, whose zeros and extrema
deliver near-optimal interpolation points for collocation methods, and
Legendre polynomials, whose orthogonality under the unit weight function
makes them the natural basis for the DG weak form.  The Legendre--Gauss--Lobatto
(LGL) nodes that the codebase uses sit at the intersection of these two
threads --- they inherit the endpoint clustering that tames the Runge
phenomenon and the quadrature exactness that makes mass lumping diagonal.

The chapter assembles the LGL toolbox piece by piece: interpolation and
the Lebesgue constant (\cref{sec:spec-interpolation}), Chebyshev polynomial
theory (\cref{sec:spec-chebyshev}), spectral differentiation matrices
(\cref{sec:spec-diffmat}), Legendre polynomials and Gaussian quadrature
(\cref{sec:spec-legendre}), convergence rates tied to function smoothness
(\cref{sec:spec-convergence}), the LGL differentiation and mass matrices
used in the DG weak form (\cref{sec:spec-lgl}), and the Vandermonde
transform that converts between nodal and modal representations for spectral
filtering (\cref{sec:spec-vandermonde}).

% ---- Drafted sections ----
\input{part3_numerical_methods/ch10_spectral_methods/spec-interpolation}
\input{part3_numerical_methods/ch10_spectral_methods/spec-chebyshev}
\input{part3_numerical_methods/ch10_spectral_methods/spec-diffmat}
\input{part3_numerical_methods/ch10_spectral_methods/spec-legendre}
\input{part3_numerical_methods/ch10_spectral_methods/spec-convergence}
\input{part3_numerical_methods/ch10_spectral_methods/spec-lgl}
\input{part3_numerical_methods/ch10_spectral_methods/spec-vandermonde}

\paragraph{Key References.}
The spectral-methods framework follows \cite{Hesthaven2007} and
\cite{Kopriva2009}; Chebyshev theory draws on \cite{Trefethen2000};
the DG application follows \cite{HesthavenWarburton2008}.
