% ch17 — Ideal MHD
\chapter{Ideal MHD}
\label{ch:ideal-mhd}

Before coupling to curved spacetime, we develop the equations and numerical
methods of ideal magnetohydrodynamics in flat space. This chapter covers the
conservative formulation, Riemann solvers, reconstruction, and the
conservative-to-primitive inversion that dominates the computational cost.

\section{Conservation Laws of Ideal MHD}
\label{sec:mhd-conservation}
\coderef{mhd}{rs}

We write the ideal MHD system in conservation form $\partial_t \mathbf{U} +
\partial_i \mathbf{F}^i = 0$, identifying the conserved density, momentum,
energy, and magnetic field variables.

\section{Primitive and Conservative Variables}
\label{sec:mhd-variables}

We define the primitive variables $(\rho, v^i, p, B^i)$ and the conservative
variables $(\mathcal{D}, S_i, \tau, B^i)$, and explain why the evolution
proceeds in conservative form while physics is expressed in primitives.

\section{Equation of State}
\label{sec:mhd-eos}
\coderef{eos}{rs}

We implement the ideal-gas EOS $p = (\Gamma - 1)\rho\epsilon$ and discuss
tabulated EOS extensions for neutron-star applications.

\section{HLL Riemann Solver}
\label{sec:mhd-hll}
\coderef{hll}{rs}

We derive the Harten-Lax-van Leer (HLL) approximate Riemann solver, estimate
the fastest wave speeds, and compute the inter-cell flux.

\section{PLM Reconstruction}
\label{sec:mhd-reconstruction}
\coderef{reconstruct}{rs}

We describe piecewise-linear reconstruction with slope limiters (minmod, MC)
that achieves second-order accuracy while suppressing spurious oscillations.

\section{Conservative-to-Primitive Inversion}
\label{sec:mhd-con2prim}
\coderef{con2prim}{rs}

We detail the Newton-Raphson root-finding procedure that recovers primitive
variables from conserved ones, including the robust handling of failures in
low-density regions.

\section{Divergence Cleaning}
\label{sec:mhd-divclean}

We introduce the hyperbolic divergence-cleaning technique of Dedner et al.\
that controls the $\nabla \cdot \mathbf{B} = 0$ constraint numerically.

\paragraph{Key References.}
The conservative formulation follows \cite{Font2008}; the HLL solver follows
\cite{Harten1983}; divergence cleaning follows \cite{Dedner2002};
reconstruction follows \cite{vanLeer1979,Toro2009}.
