% ch06 — Geodesic Motion
\chapter{Geodesic Motion}
\label{ch:geodesics}

The metrics of \cref{ch:schwarzschild,ch:kerr} define the stage; geodesics
are the actors.  Every pixel in a ray-traced black hole image records the
fate of a single null geodesic --- captured by the horizon, escaped to the
celestial sphere, or intercepted by the accretion disc --- so the quality
of the image is determined entirely by how accurately and efficiently those
geodesics are integrated.  This chapter develops the mathematical framework,
numerical machinery, and physical analysis that connect the metric to the
rendered pixel.

Two formulations of the geodesic equation set the stage.  The Lagrangian
form, built from Christoffel symbols, is the standard textbook route; the
Hamiltonian form, which replaces the velocity with the covariant momentum,
avoids differentiating the inverse metric and gives the codebase its
eight-component ODE.  Both formulations yield the same trajectories, but
the Hamiltonian version carries a built-in accuracy diagnostic: the
super-Hamiltonian $\mathcal{H}$ is conserved along every geodesic, so any
numerical drift in $\mathcal{H}$ directly measures accumulated error.
The constants of motion --- energy $E$, angular momentum $L$, and the
Carter constant $\mathcal{Q}$ for Kerr --- then reduce the system's
effective dimensionality and provide additional monitoring quantities.
Specialising to null geodesics fixes the constraint $\mathcal{H} = 0$
and introduces the affine rescaling freedom that lets the ray tracer set
$E = 1$ without loss of generality; timelike geodesics supply the disc
physics --- ISCO location, circular-orbit four-velocities, and the
Doppler shifts that create the characteristic brightness asymmetry in
accretion disc images.

The numerical core of the chapter is an adaptive Dormand--Prince integrator
whose step size shrinks by more than an order of magnitude near the photon
sphere and recovers in the weak-field escape phase --- a behaviour that
fixed-step symplectic methods cannot match without wasting computation.
Conservation monitoring tracks $\mathcal{H}$, $E$, and $L$ at every step,
providing a per-ray error certificate.  Termination conditions classify
each ray's outcome --- horizon capture, celestial-sphere escape,
disc intersection, or computational safety net --- and feed directly into
the rendering pipeline.  The chapter closes with a starred section on the
photon ring: the infinite sequence of exponentially narrowing sub-rings
that encodes the Lyapunov instability of the photon sphere and constitutes
one of general relativity's cleanest predictions about observable
black hole images.

% ---- Drafted sections ----
\input{part2_black_holes/ch06_geodesics/geo-formulations}

\input{part2_black_holes/ch06_geodesics/geo-constants}

\input{part2_black_holes/ch06_geodesics/geo-null}

\input{part2_black_holes/ch06_geodesics/geo-timelike}

\input{part2_black_holes/ch06_geodesics/geo-raytracing}

\input{part2_black_holes/ch06_geodesics/geo-conservation}

\input{part2_black_holes/ch06_geodesics/geo-termination}

\input{part2_black_holes/ch06_geodesics/geo-photon-rings}
