% ch07 — Finite Difference Methods
\chapter{Finite Difference Methods}
\label{ch:finite-differences}

Finite differences are the simplest tool for discretising derivatives on a
grid. This chapter derives the standard stencils used in the codebase,
analyses their truncation error, introduces Kreiss-Oliger dissipation for
stability, and sets up the convergence-testing framework applied throughout
the numerical-methods chapters.

\section{Finite Difference Stencil Derivation}
\label{sec:fd-stencils}
\coderef{stencils}{rs}

We derive centred finite difference stencils from Taylor expansion and show
how the codebase generates them for arbitrary order.

\section{Second- and Fourth-Order Differences}
\label{sec:fd-orders}

We write out the explicit second- and fourth-order stencils for first and
second derivatives, comparing their accuracy per grid point.

\section{Truncation Error Analysis}
\label{sec:fd-truncation}

We define the local truncation error, relate it to the stencil order, and
show how Richardson extrapolation can improve accuracy.

\section{Kreiss-Oliger Dissipation}
\label{sec:fd-dissipation}
\coderef{dissipation}{rs}

We introduce Kreiss-Oliger artificial dissipation as a mechanism to damp
high-frequency grid noise without degrading the convergence order of the
scheme.

\section{Convergence Testing}
\label{sec:fd-convergence}

We describe the method of manufactured solutions and self-convergence tests,
establishing the verification methodology used in later chapters.

\paragraph{Key References.}
Stencil derivation and error analysis follow \cite{LeVeque2007};
Kreiss-Oliger dissipation is discussed in \cite{Gustafsson1995};
convergence testing follows \cite{Roache1998}.
