% ch01 — Special Relativity: A Working Primer
\chapter{Special Relativity: A Working Primer}
\label{ch:special-relativity}

Everything in this book---the geodesic integrator, the metric functions, the
camera model, the accretion disc renderer---rests on a handful of structures
from special relativity: four-vectors, the Minkowski metric, and the causal
distinction between timelike, null, and spacelike directions.  This
chapter builds those structures from first principles, establishing the index
notation and conventions that we'll use without further ceremony for the
remaining seven hundred-odd pages.

The emphasis is deliberately computational.  We won't trace the historical
path from Michelson--Morley through Einstein's postulates to the Lorentz
transformations---excellent accounts exist elsewhere
\cite{Carroll2004,MTW1973}.  Instead, we'll work in geometrised units ($c = G
= 1$) from the start and focus on the tensorial machinery that appears
directly in the codebase: the metric as a matrix, index operations as linear
algebra, and null vectors as the starting data for ray tracing.
\histnote{Einstein's 1905 paper derived the Lorentz transformations from two
postulates.  We take the geometric view instead: define the metric, and let the
symmetries follow.}

The three sections trace a path from spacetime geometry to the camera's
viewfinder.  We begin with the Minkowski metric and the classification of
vectors into timelike, null, and spacelike --- the vocabulary that the geodesic
integrator uses to distinguish massive particles from photons.  Lorentz
transformations then enter not as abstract symmetries but through the tetrad
formalism, which constructs the observer's local reference frame from four
orthonormal vectors.  The chapter closes with the light cone and null vectors,
where we derive the explicit formula that maps a pixel on the rendered image to
a photon four-momentum --- the initial data for every ray trace.

Readers comfortable with special relativity may wish to skim this chapter for
notation and proceed to \cref{ch:differential-geometry}, where we generalise
to curved spacetime.  But the tetrad construction in \cref{sec:sr-lorentz} and
the null-vector parametrisation in \cref{sec:sr-causal} reappear verbatim in
the camera model of \cref{ch:camera-rendering}, so they're worth a careful
look even for the experienced reader.

% ---- Drafted sections ----
\input{part1_foundations/ch01_special_relativity/sr-minkowski}

\input{part1_foundations/ch01_special_relativity/sr-lorentz}

\input{part1_foundations/ch01_special_relativity/sr-causal}

\paragraph{Key References.}
The index-notation conventions follow \cite{MTW1973}; the tetrad formalism
draws on \cite{Wald1984}. The camera-tetrad construction is described in
\cite{James2015}.
