% ch09 — Automatic Differentiation
\chapter{Automatic Differentiation}
\label{ch:autodiff}

The integrator built in the previous chapter advances geodesics through
spacetime --- but at every step it needs to know how the metric changes from
point to point.  Specifically, the Hamiltonian right-hand side requires the
40 independent derivatives $\pd{\mu} g_{\alpha\beta}$: ten symmetric metric
components differentiated with respect to each of four coordinates.  For a
Schwarzschild black hole in Kerr--Schild coordinates, each derivative
involves products of rational functions and square roots in Cartesian
coordinates.  For Kerr, add trigonometric terms and a spin parameter that
threads through every expression.  Hand-deriving and maintaining these 40
expressions per spacetime is a reliable source of sign errors, transcription
bugs, and debugging sessions that end with a misplaced minus sign in the
$\christoffel{r}{\theta}{\phi}$ component.

Automatic differentiation (AD) eliminates this class of errors by
redefining arithmetic itself.  The developer writes the metric function
once --- ten numbers from four coordinates --- and the library produces
exact derivatives mechanically, with no finite-difference approximation
and no hand-coded expressions.  The key insight is algebraic: extending
the real numbers by a nilpotent element $\varepsilon$ with
$\varepsilon^2 = 0$ creates \emph{dual numbers}, and evaluating any smooth
function in this extended arithmetic yields the function's derivative in the
$\varepsilon$-coefficient, carried alongside the value at no conceptual cost.
The technique is exact to machine precision, polymorphic over spacetimes,
and cheaper than the alternatives.

The chapter develops forward-mode AD from the dual-number
construction (\cref{sec:ad-forward}), shows how the codebase
connects metric derivatives to the equations of motion via the
covariant momentum formulation (\cref{sec:ad-christoffels}),
and quantifies the accuracy advantage over finite differences
(\cref{sec:ad-comparison}).  A final optional section surveys
reverse-mode AD and its potential applications to gradient-based
parameter estimation (\cref{sec:ad-reverse}).

% ---- Drafted sections ----
\input{part3_numerical_methods/ch09_automatic_differentiation/ad-forward}

\input{part3_numerical_methods/ch09_automatic_differentiation/ad-christoffels}

\input{part3_numerical_methods/ch09_automatic_differentiation/ad-comparison}

\input{part3_numerical_methods/ch09_automatic_differentiation/ad-reverse}
