% ch04 — The Schwarzschild Solution
\chapter{The Schwarzschild Solution}
\label{ch:schwarzschild}

\Cref{sec:gr-exact-solutions} presented the Schwarzschild line element as a
finished product --- a metric and a type class instance handed to the ray
tracer without derivation.  This chapter supplies the derivation.  Spherical
symmetry and the vacuum field equations are so constraining that they
determine the metric up to a single parameter, the mass~$M$; Birkhoff's
theorem sharpens this by showing that staticity need not be assumed.  The
result is the unique exterior geometry of any spherically symmetric
gravitating body, whether a collapsing star or a static black hole.

The derived line element has a troublesome feature: its components diverge at
$r = 2M$.  Three coordinate systems repair this singularity, each offering a
different computational advantage.  Ingoing Eddington--Finkelstein coordinates
replace Schwarzschild time with a null coordinate that remains finite at the
horizon, revealing $r = 2M$ as a one-way membrane rather than a barrier.  The
Kerr--Schild form recasts the full metric as flat Minkowski space plus a
rank-one null correction --- giving the ray tracer a Cartesian,
horizon-penetrating chart whose inverse costs nothing more than a sign flip.
Kruskal--Szekeres coordinates then expose the complete causal structure,
including the white hole and second asymptotic region that Schwarzschild
coordinates conceal entirely.

The chapter closes with an effective-potential analysis of geodesic orbits.
Three characteristic radii --- the event horizon at $r = 2M$, the photon
sphere at $r = 3M$, and the innermost stable circular orbit at $r = 6M$ ---
emerge as critical points of this potential, and each serves as a termination
or boundary condition in the geodesic integrator.  Every result here
generalises to spinning black holes; the non-rotating case establishes the
patterns that \cref{ch:kerr} extends.

% ---- Drafted sections ----
\input{part2_black_holes/ch04_schwarzschild/schw-derivation}

\input{part2_black_holes/ch04_schwarzschild/schw-coordinates}

\input{part2_black_holes/ch04_schwarzschild/schw-kerr-schild}

\input{part2_black_holes/ch04_schwarzschild/schw-kruskal}

\input{part2_black_holes/ch04_schwarzschild/schw-orbits}

\paragraph{Key References.}
The derivation follows \cite{Carroll2004}; Kerr-Schild coordinates are
discussed in \cite{Visser2007}; the orbital analysis follows \cite{Wald1984}.
